A multinomial spatio-temporal model for sexual risk behaviour in AGYW

### 2021-08-09 docs/09-09-21_three-category/three-category.Rmd
Slides up

---
title: A multinomial spatio-temporal model for sexual risk behaviour in adolescent girls and young women
subtitle: HIV Inference Lab Group Meeting
author: Adam Howes
institute: Imperial College London
date: August 2021
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

@@

* Want to estimate the proportion $p_k$ of AGYW in $K = 4$ risk categories
  * $k = 1$: Not sexually active
  * $k = 2$: Cohabiting, single partner
  * $k = 3$: Non-regular partner(s)
  * $k = 4$: Key population, female sex worker
* Used as a step towards estimating total incidence by group
  * By district, age, sex, time
* Different policy responses depending on where new infections are

## Extent

* 13 AGYW Global Fund priority countries (but fitting individually)
  * Botswana, Cameroon, Kenya, Lesotho, Malawi, Mozambique, Namibia, South Africa, Swaziland, Tanzania, Uganda, Zambia and Zimbabwe
* The available DHSs

@@

* Estimate proportions indexed by
  * District $i = 1, \ldots, n$
  * Survey $t = 1, \ldots, T$
  * Age $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$
  * Category $k$
* e.g. in Lilongwe, in 2015, how many AGYW aged 20-24 are cohabiting with a single partner?

@@

* Requires a multivariate model in that each observation depends on $K$ probabilities
  * Really $K - 1$ numbers, since they are constrained to sum to one
* Usually this is done by modelling the "contrasts" between categories
  * How much more likely to be in this category (or group of categories) than that category (or group of categories)
* These types of models are popular in economics to model choice
  * Treat the latent field like utilities of choices
* Some keywords: discrete choice model, multinomial logit model

@@

@@

@@

* $y_{itak} \sim \text{Poisson}(\lambda_{itak})$
* $\log(\lambda_{itak}) = \eta_{itak}$ the linear predictor for district $i$, survey $t$, age-group $a$ and category $k$
* Including observation-specific random effects $\theta_{ita} \sim \mathcal{N}(0, \tau_\theta^{-1})$ in the model $\eta_{itak} = \theta_{ita} + \cdots$ assures exact reproduction of the multinomial denominators
  * By "observation" I mean observation of the multinomial e.g. a vector `c(2, 5, 3, 0)`
* Then you put whatever mixed model things you'd like in

@@

>>>>

---
title: A multinomial spatio-temporal model for sexual risk behaviour in AGYW
subtitle: HIV Inference Lab Group Meeting
author: Adam Howes
institute: Imperial College London
date: August 2021
bibliography: citations.bib
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

@@

* Want to estimate the proportion $p_k$ of AGYW in $K = 4$ risk categories
  * $k = 1$: Not sexually active `nosex12m`
  * $k = 2$: Cohabiting, single partner `sexcohab`
  * $k = 3$: Non-regular partner(s) `sexnonreg`
  * $k = 4$: Key population, female sex worker `sexpaid12m`
* Used as a step towards estimating total incidence by group
  * By district, age, sex, time
* Different policy responses depending on where new infections are

@@

* Estimate proportions indexed by
  * District $i = 1, \ldots, n$
    * 13 AGYW Global Fund priority countries (but fitting individually): Botswana, Cameroon, Kenya, Lesotho, Malawi, Mozambique, Namibia, South Africa, Swaziland, Tanzania, Uganda, Zambia and Zimbabwe
  * Survey $t = 1, \ldots, T$
    * The available DHSs
  * Age $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$
  * Category $k$
* e.g. in Lilongwe, in 2015, how many AGYW aged 20-24 are cohabiting with a single partner?

## Why not just use the raw data

1. Usual high variance from low counts problem, improved by spatio-temporal smoothing
2. The raw proportions don't add up to one
  * This is illustrated in the next slide

##

:::: columns

::: column

```{r echo=FALSE}
knitr::include_graphics("page1-check100.pdf")
```

:::

::: column

```{r echo=FALSE}
knitr::include_graphics("page2-check100.pdf")
```

:::

::::

@@

* Requires a multivariate model in that each observation depends on $K$ probabilities
  * Really $K - 1$ numbers, since they are constrained to sum to one
* Usually this is done by modelling the "contrasts" between categories
  * How much more likely to be in this category (or group of categories) than that category (or group of categories)
  * Different ways to do this
* These types of models are popular in economics to model choice
  * Treat the latent field like utility
* Some keywords: discrete choice model, multinomial logit model

@@

## Sidenote: ordinal

* $\pi_{k} = \sum_{l \leq k} p_{l}$ be the cumulative probability of category $k$ where $\pi_{1} \leq \pi_{2} \leq \cdots \leq \pi_{K}$.
* Ordinal logistic regression is based upon $K - 1$ cumulative logits $\text{logit} (\pi_{k}) = \eta_{k}$ where the linear predictors are ordered
* One way to do this is to assume $\eta_k$ only differ in their intercept (also called cut-point)
* I wonder if you can fit this with a single linear predictor and a category random effect?
  * Possibly wouldn't need Poisson trick

@@

* "The Poisson surrogate model eliminates $m$ from the denominator of the multinomial probabilities. This makes sense intuitively, as we do not expect the multinomial sums to provide any useful information in estimating the fixed effects." [@lee2017poisson]
* Example of a broader class of Poisson transformations where the theme is to consider the normalisation constant as another parameter [@barthelme2015poisson]
* I guess that this must mean the normalisation constant is "estimated" exactly always

@@

* $y_{itak} \sim \text{Poisson}(\lambda_{itak})$
* $\log(\lambda_{itak}) = \eta_{itak}$ the linear predictor for district $i$, survey $t$, age-group $a$ and category $k$
* Including observation-specific random effects $\theta_{ita} \sim \mathcal{N}(0, \tau_\theta^{-1})$ in the model $\eta_{itak} = \theta_{ita} + \cdots$ assures exact reproduction of the multinomial denominators
  * By "observation" I mean observation of the multinomial e.g. a vector `c(2, 5, 3, 0)`
  * These variables correspond to $\log(\delta)$ from before. So we're just having a free parameter which can 
* Then you put whatever mixed model things you'd like in

@@

## Survey weights

* Use survey weights to calculate effective observation `x_eff`
* May not be integers i.e. `c(1.87, 5.32, 2.20, 0.15)`
* Deal with this by using `xPoisson` (generalisation of the Poisson to non-integer data) rather than `Poisson`
* Might be equivalent to weighted log-likelihood approach

##

:::: columns

::: column

```{r echo=FALSE}
knitr::include_graphics("page-multinomial-smoothed-district-sexbehav.pdf")
```

:::

::: column

* A (difficult to see) example of what this looks like
* Model 9 has BYM2 and AR1
* Other notes
  * Quantiles produced by manipulating `inla.posterior.sample`s
  * Same samples used in combination with Naomi population data to produce aggregates (i.e. 15-24 age category)

:::

::::

## Data issues for `sexpaid12m` category

* Two different questions asked in the DHS, and answered very differently
* Before 2015, asked about most recent three partners. One of the categories for these is paid
  * Very rare to list one of these as paid, using this question often think there are essentially no FSW
* After 2015, asked if have given or received gifts or money in exchange for sex
  * Better answered but still has issues
    * Missing some FSW from survey sampling frame
    * Not all "gifts of money in exchange for sex" is sex work

##

```{r echo=FALSE}
knitr::include_graphics("../../archive/check_sexpaid-survey-question/20210720-120340-77026581/sexpaid-survey-question.pdf")
```

## Current approach

* Can't fit a (meaningful) spatio-temporal model when the question asked / response is so different
* Have been fitting single-survey spatial models to the most recent DHS
* Otherwise have created a three-category model, moving all `sexpaid12m` into `sexnonreg` to create a new `sexnonregplus` category
  * Haven't checked that some of the `sexpaid12m` shouldn't go into `sexcohab` rather than `sexnonreg` yet (sorry Katie!)
* Challenge is what to do about the `sexpaid12m` category

## What to do about the `sexpaid12m` category

1. Ignore it
2. Try to use other data sources to help
  * Previous estimates from workbook are based upon national estimates of FSW population size (perhaps the UNAIDS Key Population Atlas)

## Johnston et al. (2021, preprint)

**Deriving and interpreting population size estimates for adolescent and young key populations at higher risk of HIV transmission: men who have sex with men, female sex workers and transgender persons**

* Disaggregates the UNAIDS published population size estimates by age (nice, but not by district, which we'd also want) using proportion of sexually active adults
* Kinh is a coauthor and warns that
  * The estimates should (often) be seen as expert opinion rather than based on data
  * Several countries had no data
  * Rounding up when the number is too small

## Laga et al. (2021, preprint)

**Mapping female sex worker prevalence (aged 15-49 years) in sub-Saharan Africa**

* Collates available FSW population size estimates
* Small-area model with covariates used to extrapolate to district level
* Has code available
* Probably the best single source to rely on, done a lot of the work already

## Hodgkins et al. (2021, preprint)

**HIV prevalence, population sizes, and HIV prevention among men who paid for sex in sub-Saharan Africa: a meta-analysis of 82 population-based surveys (2000-2020)**

* Proportion of men who pay for sex (CFSW) estimated from survey data
* Could be linked to proportion FSW by some model e.g.
$$
p_{\text{CFSW}} = B \times p_{\text{FSW}}
$$
with a strongly informative prior $B \sim p(B)$ (around 10 say)
  * Don't know if this is a good model, anyone have any ideas?
  * Extract estimates from Hodgkins and plot versus estimates of FSW to see if it looks linear
  * Of course the ratio is going to vary, but perhaps it's quite stable?
* Fit a model on the good DHS `sexpaid12m` question, then use it where there isn't the question

##

\begin{align}
\log(p_{\text{CFSW}}) - \log(p_{\text{FSW}}) &= \log(B) \\
\log \left( \frac{p_{\text{CFSW}}}{p_{\text{FSW}}} \right) &= \log(B)
\end{align}
